---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# virgo

<!-- badges: start -->
<!-- badges: end -->

The **virgo** package enables the creation of interactive graphics for
exploratory data analysis. It is an _idiomatic and opinionated_ R interface to 
the grammar of graphics implemented by [**Vega-Lite**](https://vega.github.io/vega-lite/) which
defines the following elements:

* graphical elements like `mark_point()`, with the `mark_*` family of functions
* interactive elements, such as brushes (using `select_interval()`) and sliders (using `input_slider()`),  via the `select_*` and `input_*` family of functions
* aesthetic mappings/encodings via `enc()` 
* data transformations via [selections]()
* plot composition via faceting and concatenation using `facet_views()`, `hconcat()` and `vconcat()`


## Installation

<!--
You can install the released version of **virgo** from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("virgo")
```

-->

You can install the development version of **virgo** 
from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("vegawidget/virgo")
```

## Get started

For most graphics using **virgo**, you start off by passing data to 
the `vega()` function, add graphical elements with marks like `mark_point()`, 
and specify variables within a mark using encodings `enc()`. You can add more 
layers by specifying additional marks like `mark_smooth()`, or include small 
multiples with `facet_views()` or combine plots or 
add interactive elements with selections.

Let's see an example, here we show how we can compose a simple scatter plot
and gradually build up to a scatter plot with brushing, to a side by side scatter plot.

```{r basic-scatter}
library(virgo)
library(palmerpenguins)
p <- penguins %>% 
  vega() %>% 
  mark_circle(
    enc(
      x = bill_length_mm, 
      y = bill_depth_mm
    )
  )
p
```

Interactive elements are generated using selections, for example, 
we can generate a rectangular brush with `select_interval()` and 
then highlight points that fall into the brush using `encode_if()`:

```{r brushed-scatter}
selection <- select_interval()

p <- penguins %>% 
  vega() %>% 
  mark_circle(
    enc(
      x = bill_length_mm, 
      y = bill_depth_mm, 
      color = encode_if(selection, species, "black")
    )
  )
p
```


Now we generate a second scatter plot:

```{r right-scatter}
p_right <- penguins %>% 
  vega() %>% 
  mark_circle(
    enc(
      x = flipper_length_mm,
      y = body_mass_g,
      color = encode_if(selection, species, "black")
    )
  )
p_right
```

And link them side by side using concatenation.

```{r linked-brushed-scatter}
hconcat(p, p_right)
```

And that's just the beginning of what's possible!

## Learning more

- [Example gallery]()
- [Using **virgo** to explore Melbourne's microclimate]()
- [Guide to **virgo** for **ggplot2** users]()
- [Composing plot interactions with selections]()

## Lifecycle

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

The **virgo** package is under rapid development and we are still working 
through our ideas for incorporating interactive graphics
into exploratory data analysis. If you have feedback we would love to hear it!

## Acknowledgements

- Vega/Vega-Lite developers 
- Ian Lyttle, Hayley Jepson and Alicia Schep for their foundational
work in the [**vegawidget**](https://vegawidget.github.io/vegawidget/) package

