---
title: "Introducing virgo for interactive exploratory graphics"
author: "Stuart Lee and Earo Wang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing virgo for interactive exploratory graphics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# About virgo

The **virgo** package extends the grammar of interactive graphics based on the **Vega-Lite** library in Javascript, using the design principles of the **tidyverse** suite of packages. It is built on top of the **vegawidget** package, which handles the actual drawing of a **virgo** graphic. 

## Melbourne's microclimate

Throughout this vignette we will be exploring three months of microclimate sensor data collected from December 2019 until February 2020 from Melbourne City Council. This data is built into to **virgo** and consists of measurements for things like temperature and relative humidity at 5 different locations across the city. 
```{r setup, echo = TRUE}
library(virgo)
library(lubridate)
library(dplyr)
melbweather
```

## Data, Encodings and Marks

To construct a visualisation, we begin with tidy data and map columns to visual elements using encodings:

```{r}
melbweather %>%
  filter(date == "2019-12-01") %>% 
  vega(encoding = enc(x = ambient_temperature))
```

The data is piped into the `vega()` function and aesthetic mappings are
specified with the `enc()`. Valid encodings depend on the **mark** or graphical element being used. As there are no marks for this **virgo** chart, blank canvas is printed. The `vega()` function is required to be called in order to produce a valid visualisation, however as well we see later both data and marks can be specified elsewhere.

To produce interesting charts, we can add **marks** which are visual layers that are added to the chart. 

Let's go back to our data:

```{r}
melbweather
```

To begin, we will collapse our quarter hourly 
measurements for each site to hourly averages using **dplyr**:

```{r, message = FALSE}
hourly_weather <- melbweather %>% 
  group_by(site, date, hour_of_day = hour(date_time) ) %>% 
  summarise(
    across(
      c(ambient_temperature:pm10), 
      mean, 
      na.rm = TRUE
    )
  ) %>% 
  ungroup()

hourly_weather
```
Suppose we are interested in the relationship between ambient temperature and relative humidity. It is theorised that as ambient temperature increases, relative humidity decreases.

We can check whether this is the case with our data by creating a scatter plot.

```{r}
hourly_weather %>% 
  vega() %>% 
  mark_point(enc(x = ambient_temperature, y = relative_humidity))
```

For users, of **ggplot2** the above incantation should look somewhat familiar. We specify our data at the top level using `vega()` and then add layers using the pipe `%>% ` in combination with marks. Inside the `mark_point()` function we specify the aesthetic mapping or **encoding** that says the x axis should correspond to ambient temperature and the y axis should correspond to relative humidity.

The **virgo** library will automatically add legends depending on the type of encoding and input variable. If we color by a continuous variable a scale is automatically determined and placed on the right hand side of the plot. 

```{r}
hourly_weather %>% 
  vega() %>% 
  mark_point(
    enc(
      x = ambient_temperature, 
      y = relative_humidity, 
      color = wind_speed
    )
  )
```

Likewise, we can evaluate expressions inside of the encoding function, for example, evaluating the month of the date column
with **lubridate**:

```{r}
hourly_weather %>% 
  vega() %>% 
  mark_point(
    enc(
      x = ambient_temperature, 
      y = relative_humidity, 
      color = month(date, label = TRUE)
    )
  )
```


It appears that there is a slight non-linear and negative relationship between ambient temperature and relative humidity, however, there is a lot over-plotting in the first example, so perhaps it would be best to add some opacity to the previous scatter plot. Graphical elements that don't depend on columns of the data can be added via extra arguments to the mark function:


```{r}
hourly_weather %>% 
  vega() %>% 
  mark_point(
    enc(x = ambient_temperature, y = relative_humidity),
    opacity = 0.1
  )
```

Multiple marks can be added to a visualisation by piping a sequence of marks together. We can take our previous scatter plot and draw a loess regression line on top.

```{r}
p <- hourly_weather %>% 
  vega(enc(x = ambient_temperature, y = relative_humidity)) %>% 
  mark_point(opacity = 0.1) %>% 
  mark_smooth(method = "loess", color = "blue")
p
```

Here we've specified the encoding at the top level with the `vega()` function, so they are available to all downstream layers. We can have layer specific encodings, by passing the `enc()` to any mark in the sequence. In addition to a single regression line, we could generate separate lines for each measurement within each month of the year:

```{r}
p %>% 
  mark_smooth(
    enc(color = month(date, label = TRUE)),
    method = "loess"
  )
```

## Facets

Sometimes we want to visualise multiple categorical variables at the same time, by creating small multiples (facets) views. Say we were interested in the distribution of wind speed, we could
create a histogram:

```{r}
ws <- hourly_weather %>% 
  vega() %>% 
  mark_histogram(enc(x = wind_speed), bin = list(step = 1))
ws
```

How similar are the wind speeds across each measurement site? We can facet by site to find out:

```{r}
ws %>% 
  facet_views(row = site)
```

Faceting requires specifying a row and/or column encoding, that determines how the subplots are drawn on the canvas.
There appears to be a suspiciously large number of records in site "arc1045" that have recorded an average wind speed between 4-5km. 


## Multiple Views

Often it is helpful to place views side by side that shows all
of the data rather than subplots like those produced by faceting.

**virgo** provides functions to align plots horizontally or vertically with the `hconcat()` or `vconcat()` functions.

```{r}
left <- hourly_weather %>% 
  vega() %>% 
  mark_point(enc(x = ambient_temperature, y = relative_humidity),
             opacity = 0.1)
right <- hourly_weather %>%
  vega() %>% 
  mark_point(enc(x = wind_speed, y = ambient_temperature),
             opacity = 0.1)
hconcat(left, right)
```


## Client side data transformations

The **virgo** package exports special functions that are computed directly by **Vega-Lite** rather than R. These functions are prefixed with `vg`, and are always be called inside of `enc()`.


By using the `vg` functions, we can get **Vega-Lite** to perform
data transformations or aggregations without the use of **tidyverse**. Suppose we were interested in finding out the hour of the day, where the 
average temperature across all sites reaches its max:
 
```{r}
hourly_temp <- melbweather %>% 
  vega(
    enc(
      x = vg_hours(date_time), 
      y = vg_mean(ambient_temperature)
    )
  ) %>% 
  mark_point() %>% 
  mark_line()
hourly_temp
```

In the above code, the `vg_hours()` function extract the hour from the date time variable and groups the data within each our block. The y-axis is then average ambient temperature within each hour pooled across all days and sites. We could get a more granular view by using an alternative mark like a boxplot to view the distribution over each hour instead of just the average:

```{r}
hourly_dist_temp <- melbweather %>% 
  vega(
    enc(
      x = vg_hours(date_time), 
      y = ambient_temperature
    ),
    width = 600
  ) %>% 
  mark_boxplot()
hourly_dist_temp
```

Generally, we believe it is best to perform data transformations outside of the visualisation environment as the transformations are explicit (i.e. when using the **tidyverse**). However, as we will see later these client side transformations become especially useful when combined with interactivity.

## Interactivity via selections

Selections are how a **virgo** graphic defines interactions, and are strongly influenced by the **Vega-Lite** javascript API. There are several types of selection, but to begin we will create an interval selection:

```{r}
selection <- select_interval()
```

By default, an interval selection will specify a rectangular region, that is generated by dragging the mouse over the view. Each mark can accept a selection object, which will filter the data when the region is active:

```{r}
p <- hourly_weather %>% 
  vega(enc(x = ambient_temperature, y = relative_humidity)) %>% 
  mark_point(opacity = 0.1, selection = selection) 
p
```

If you just want to draw the selection without performing the filter, you can specify it as an identity selection

```{r}
p <- hourly_weather %>% 
  vega(enc(x = ambient_temperature, y = relative_humidity)) %>% 
  mark_point(opacity = 0.1, selection = I(selection)) 
p
```

Rather than filtering the data, you may want to conditionally encode a visual element based on whether data falls into a selection. The `encode_if()` function allows an encoding to depend on one (or more) selections. We could rewrite the above plot so the points that fall inside of the selection are given an opacity of 0.5, while those outside are given an opacity of 0.1.

```{r}
p <- hourly_weather %>% 
  vega(enc(x = ambient_temperature, y = relative_humidity)) %>% 
  mark_point(
    enc(opacity = encode_if(selection, 0.5, 0.1))
  ) 
p
```

Instead of using values, the conditional encoding can depend on a variable instead:

```{r}
p <- hourly_weather %>% 
  vega(enc(x = ambient_temperature, y = relative_humidity)) %>% 
  mark_point(
    enc(
      color = encode_if(selection, wind_speed > 10, "grey"),
      opacity = encode_if(selection, 0.5, 0.01)
      )
  ) 
p
```

Interval selections can be restricted to a single axis, so a brush only moves in one direction. The following produces a line plot of the maximum daily temperature, with a brush over the x-axis

```{r}
x_selection <- select_interval(encodings = "x")

overview <- melbweather %>% 
  vega(
    enc(
      x = date,
      y = vg_max(ambient_temperature)
    ),
    width = 600
  ) %>% 
  mark_line() %>% 
  mark_point(
    selection = I(x_selection)
  )
overview
```

Interval selections can also be bound to a scale and then composed with another
view to produce a zooming effect.
After concatenating the two views, dragging the x-axis on the bottom graphic will zoom in on the top view:

```{r}
context <- melbweather %>% 
  vega(
    enc(
      x = date,
      y = vg_max(ambient_temperature)
    ),
    width = 600,
    height = 200
  ) %>% 
  mark_line() %>% 
  mark_point() %>% 
  scale_x(name = NULL, domain = x_selection)

vconcat(context, overview)
```

Selections can also be used to filter data in multi-view plots.  For example we could view the trail of temperature measurements that each site recorded over the highlighted period in the overview plot.
```{r}
active_sensor <- melbweather %>% 
  vega(enc(x = ambient_temperature, y = site), height = 200) %>% 
  mark_tick(selection = x_selection) 
vconcat(overview, active_sensor)
```

This is just the beginning of what is possible to achieve with selections, for more detail see the selections vignette.

## Sizing and saving

The sizing of a plot is controlled inside the `vega()` function with the height and width arguments. These arguments refer to the pixel dimensions of the view.

**virgo** graphics can be saved with the **vegawidget** package using the `vg_write_png()` or `vg_write_svg()` functions.
