---
title: "Model checking with filtering selections"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(virgo)
library(dplyr)
library(tidyr)
library(gapminder)
```

This example is inspired by the [Many Models](https://r4ds.had.co.nz/many-models.html#gapminder) example in R for Data Science by Grolemund and Wickham.

To begin we create a drop down menu with each continent
using `input_select()` and then we create a line plot for each country's life expectancy over time. Here the selection modifies
the opacity of each line, so we can compare countries within a continent to each other:
```{r}
selection <- select_bind(
  continent = input_select(
    name = "Select Continent:",
    choices = c(NA, levels(gapminder$continent)))
)
  
lex <- gapminder %>% 
  vega(enc(x = year, y = lifeExp, group = country)) %>% 
  mark_line(
    enc(opacity = encode_if(selection, 0.5, 0.1))
  )
lex 
```

We see that for most countries life expectancy is generally increasing, however several countries from Africa and Asia have dips in their life expectancies. 

Next for each country we can estimate a simple linear model to summarise the relationship of how life expectancy has changed over time. We will use $R^2$ to assess the quality of the fit
and link that back to our raw data

```{r}
# we could do this with broom but this seems
# ok for now
gapminder_augmented <- gapminder %>% 
    group_nest(continent, country) %>% 
    mutate(
      model = lapply(data, function(x) lm(lifeExp ~ year, data = x)),
      r2 = vapply(model, function(x) summary(x)$r.squared, numeric(1)),
    ) %>% 
  unnest(data) %>% 
  select(-model)
  
gapminder_augmented
```

Now we can create a plot driven selection, that will display the
$R^2$ values within each continent:

```{r}
select_r2 <- select_interval("y")

tick_plot <- gapminder_augmented %>% 
    vega(enc(x = continent, y = r2)) %>% 
    mark_tick(
      enc(opacity = encode_if(select_r2, 1, 0.1))
    )
tick_plot
```


New we can overlay the real data alongside the model $R^2$ values to 
identify which countries have poor fits. First, we modify the color scale
to match the provided `gapminder::country_colors` and then produce a line plot.

```{r}
palette <- country_colors[sort(names(country_colors))]

country_fits <- gapminder_augmented %>% 
  vega(enc(x = year, y = lifeExp, color = country)) %>% 
  mark_line(selection = select_r2) %>% 
  scale_color(range = palette ,guide = FALSE)

country_fits
```

Next we combine them together to see which countries have poor fits.

```{r}
hconcat(tick_plot, country_fits)
```
The countries with $R^2$ below 0.2 have strong non-linear trends in life expectancies
hovering over the lines identifies them to be countries affected by genocide
(Rwanda) or the HIV/AIDS epidemic (Botswana, Lesotho, Swaziland, Zambia, Zimbabwe).

